---
title: "SCEM Assignment 6"
author: "Max Kirwan"
date: "10/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
```


## 1. A Gaussian model for red tailed hawks
```{r}
library(Stat2Data)
data("Hawks")

RedTailedDf <- Hawks %>%
  filter(Species=="RT") %>%
  select(c("Weight","Tail","Wing"))
RedTailedDf %>% head(5)
```

```{r}
# Finding the maximum likelihood estimates for mu and sigma
RedTailedTail <- RedTailedDf %>% pull("Tail")
mu_MLE <- sum(RedTailedTail)/length(RedTailedTail)
sigma_MLE <- sqrt(sum((RedTailedTail-mu_MLE)**2)/length(RedTailedTail))

weights <- seq(mu_MLE-5*sigma_MLE,mu_MLE+5*sigma_MLE,sigma_MLE*0.001)
colours <- c("MLE Density"="red", "Kernel Density"="blue")
ggplot() + geom_line(data=data.frame(Weight=weights,Density=dnorm(weights,mu_MLE,sigma_MLE)),aes(x=Weight,y=Density,colour="MLE Density")) + geom_density(data=RedTailedDf, aes(x=Tail,colour="Kernel Density")) + labs(x="Tail Length (mm)",y="Density",colour="Estimator") + theme_bw() + scale_colour_manual(values=colours)
```

## 2. Location estimators with Gaussian data

```{r}
set.seed(0)
num_trials_per_sample_size <- 100
min_sample_size <- 5
max_sample_size <- 1000
sample_size_inc <- 5
mu_0 <- 1
sigma_0 <- 3

simulation_df <- crossing(trial=seq(num_trials_per_sample_size),sample_size=seq(min_sample_size,max_sample_size,sample_size_inc)) %>% # create data frame of all pairs of sample_size and trial
  mutate(simulation=pmap(.l=list(trial,sample_size), .f=~rnorm(.y,mean=mu_0,sd=sigma_0))) %>% # simulate sequences of Gaussian random variables
  mutate(sample_mean=map_dbl(.x=simulation,.f=mean)) %>% # compute the sample means
  mutate(sample_median=map_dbl(.x=simulation,.f=median)) %>% # compute the sample medians
  group_by(sample_size) %>%
  summarise(mse_mean=mean((sample_mean-mu_0)**2),mse_median=mean((sample_median-mu_0)**2))

simulation_df <- simulation_df %>%
  pivot_longer(cols=c(mse_mean,mse_median),names_to="estimator",values_to="mse")

ggplot(data=simulation_df,aes(x=sample_size,y=mse,colour=estimator)) + geom_smooth() + labs(x="Sample size",y="Mean squared error")
```

## 3. Unbiased estimation of the population variance

```{r}
set.seed(0)
num_trials_per_sample_size <- 100
min_sample_size <- 5
max_sample_size <- 1000
sample_size_inc <- 5
mu_0 <- 1
sigma_0 <- 3

var_MLE <- function(X,X_bar){return(sum((X-X_bar)**2)/length(X))}
var_U <- function(X,X_bar){return(sum((X-X_bar)**2)/(length(X)-1))}

simulation_df_2 <- crossing(trial=seq(num_trials_per_sample_size),sample_size=seq(min_sample_size,max_sample_size,sample_size_inc)) %>% # create data frame of all pairs of sample_size and trial
  mutate(simulation=pmap(.l=list(trial,sample_size), .f=~rnorm(.y,mean=mu_0,sd=sigma_0))) %>% # simulate sequences of Gaussian random variables
  mutate(var_MLE=map_dbl(.x=simulation,.f=~var_MLE(.x,mean(.x))), var_U=map_dbl(.x=simulation,.f=~var_U(.x,mean(.x)))) %>% # compute the two variance estimators
  mutate(sd_MLE=map_dbl(.x=simulation,.f=~sqrt(var_MLE(.x,mean(.x)))),sd_U=map_dbl(.x=simulation,.f=~sqrt(var_U(.x,mean(.x))))) %>% # compute the two sd estimators
  group_by(sample_size) %>%
  summarise(bias_var_MLE=mean(var_MLE)-sigma_0**2,bias_var_U=mean(var_U)-sigma_0**2,bias_sd_MLE=mean(sd_MLE)-sigma_0,bias_sd_U=mean(sd_U)-sigma_0)

simulation_df_2 <- simulation_df_2 %>%
  pivot_longer(cols=c(bias_var_MLE,bias_var_U,bias_sd_MLE,bias_sd_U),names_to="estimator",values_to="bias")

ggplot(data=simulation_df_2,aes(x=sample_size,y=bias,colour=estimator)) + geom_smooth() + labs(x="Sample size",y="Bias",colour="Estimator")
```

## 4. Maximum likelihood estimates for the Gaussian distribution

See hand written notes.


## 5. Maximum likelihood estimate with the Poisson distribution

See hand written notes for proof that $\hat{\lambda}_{MLE} = \bar{X}$.

```{r}
set.seed(0)
num_trials_per_sample_size <- 100
min_sample_size <- 5
max_sample_size <- 1000
sample_size_inc <- 5
lambda_0 <- 0.5

simulation_df_3 <- crossing(trial=seq(num_trials_per_sample_size),sample_size=seq(min_sample_size,max_sample_size,sample_size_inc)) %>% # create data frame of all pairs of sample_size and trial
  mutate(simulation=pmap(.l=list(trial,sample_size), .f=~rpois(.y,lambda_0))) %>% # simulate sequences of Gaussian random variables
  mutate(sample_mean=map_dbl(.x=simulation,.f=mean)) %>% # compute the sample means
  group_by(sample_size) %>%
  summarise(mse=mean((sample_mean-lambda_0)**2))

ggplot(data=simulation_df_3,aes(x=sample_size,y=mse)) + geom_smooth() + labs(x="Sample size",y="Mean squared error")
```

```{r}
VonB <- read.csv("/Users/maxkirwan/Desktop/Uni/Data Science MSc/SCEM/Statistical-Computing-and-Empirical-Methods/VonBortkiewicz.csv")

# Modelling the fatalities as independent r.v.'s from a Poisson distribution with parameter lambda
# Computing the maximum likelihood estimate for lambda
fatalities_MLE <- mean(pull(VonB,"fatalities"))
print(fatalities_MLE)

# Calculating the probability that a single cavalry corps has no fatalities in a single year
print(dpois(0,fatalities_MLE))
```


## 6. Maximum likelihood estimate for the exponential distribution

See hand written notes for proof that $\hat{\lambda}_{MLE} = \frac{1}{\bar{X}}$.

```{r}
CustomerPurchases <- read.csv("/Users/maxkirwan/Desktop/Uni/Data Science MSc/SCEM/Statistical-Computing-and-Empirical-Methods/CustomerPurchases.csv")

CustomerPurchases <- CustomerPurchases %>%
  mutate(time_diffs=lead(Time)-Time)

# Modelling the time differences as independent r.v.'s from an exponential distribution with parameter lambda
# Computing the maximum likelihood estimate for lambda
time_diffs_MLE <- 1/mean(pull(CustomerPurchases,time_diffs),na.rm=1)
print(time_diffs_MLE)

# Estimating the probability of an arrival time in excess of one minute
print(1-pexp(60,time_diffs_MLE))
```


