---
title: "SCEM Assignment 5"
author: "Max Kirwan"
date: "27/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
library(SciViews)
```

## 1. Expectation and variance of a discrete random variable

Answers written in note book.


## 2. Simulating data with the uniform distribution

Creating a sequence of i.i.d. copies $X_1$,...,$X_n$ with $\alpha = \beta = 0.25$.
```{r}
set.seed(0)
n <- 1000
sample_X <- data.frame(U=runif(n)) %>%
  mutate(X=case_when((0<=U)&(U<0.25)~1, (0.25<=U)&(U<0.5)~5, (0.5<=U)&(U<=1)~0)) %>%
  pull(X)
sample_X
```

```{r}
# Function to create a sequence of i.i.d.'s with inputs alpha, beta, n.
sample_X_015 <- function(alpha, beta, n){
  set.seed(1)
  sample_X_vec <- data.frame(U=runif(n)) %>%
    mutate(X=case_when((0<=U)&(U<alpha)~1, (alpha<=U)&(U<alpha+beta)~5, (alpha+beta<=U)&(U<=1)~0)) %>%
    pull(X)
  return(sample_X_vec)
}
```

```{r}
# Computing the sample average
mean(sample_X_015(0.5,0.1,10000))
```
The sample average is 0.9982. This is very close to the expectation, which is $\mathbb{E}(X)= \alpha + 5\beta = 0.5 + 5\times 0.1 = 1$.
This is due to the law of large numbers: the sample average converges towards the expectation.

```{r}
sample_averages <- data.frame(beta=seq(0.1,0.9,0.01)) %>%
    mutate(mean=map_dbl(beta, ~mean(sample_X_015(0.1,.x,100))))
ggplot(data=sample_averages, aes(x=beta, y=mean)) + geom_col() + xlab("Beta") + ylab("Sample Mean")
```


## 3. The Gaussian distribution

A Gaussian random variable with mean $\mu$ and standard deviation $\sigma > 0$ has probability density function $f(x) = \frac{1}{\sigma \sqrt{2\pi}}e^{-\frac{(x-\mu )^2}{2\sigma ^2}}$.

```{r}
grid <- seq(-4,6,0.01)
normal_distributions <- data.frame(grid) %>%
  mutate(one=map_dbl(grid,~dnorm(.x,1,1))) %>%
  mutate(two=map_dbl(grid,~dnorm(.x,1,sqrt(2)))) %>%
  mutate(three=map_dbl(grid,~dnorm(.x,1,sqrt(3))))
ggplot(data=normal_distributions, aes(x=grid)) + geom_line(aes(y=one,colour="red")) + geom_line(aes(y=two,colour="blue")) + geom_line(aes(y=three,colour="green")) + labs(title="Probability density function",x="x",y="Density",colour="Variance") + scale_color_manual(labels = c("1","2","3"), values = c("red","blue","green"))
```

```{r}
grid <- seq(-4,6,0.01)
normal_distributions <- data.frame(grid) %>%
  mutate(one=map_dbl(grid,~pnorm(.x,1,1))) %>%
  mutate(two=map_dbl(grid,~pnorm(.x,1,sqrt(2)))) %>%
  mutate(three=map_dbl(grid,~pnorm(.x,1,sqrt(3))))
ggplot(data=normal_distributions, aes(x=grid)) + geom_line(aes(y=one,colour="red")) + geom_line(aes(y=two,colour="blue")) + geom_line(aes(y=three,colour="green")) + labs(title="Cumulative distribution function",x="x",y="Probability",colour="Variance") + scale_color_manual(labels = c("1","2","3"), values = c("red","blue","green"))
```

```{r}
grid <- seq(0,1,0.01)
normal_distributions <- data.frame(grid) %>%
  mutate(one=map_dbl(grid,~qnorm(.x,1,1))) %>%
  mutate(two=map_dbl(grid,~qnorm(.x,1,sqrt(2)))) %>%
  mutate(three=map_dbl(grid,~qnorm(.x,1,sqrt(3))))
ggplot(data=normal_distributions, aes(x=grid)) + geom_line(aes(y=one,colour="red")) + geom_line(aes(y=two,colour="blue")) + geom_line(aes(y=three,colour="green")) + labs(title="Quantile function",x="Quantile",y="x",colour="Variance") + scale_color_manual(labels = c("1","2","3"), values = c("red","blue","green"))
```


## 4. The Binomial distribution and the central limit theorem

```{r}
binom_df <- data.frame(x=0:50) %>%
  mutate(pmf=map_dbl(x,~dbinom(.x,50,0.7)))
binom_df %>% head(3)
```
```{r}
mu <- 50*0.7
sigma <- sqrt(mu*(1-0.7))
gaussian_df <- data.frame(x=seq(0,50,0.01)) %>%
  mutate(pdf=map_dbl(x,~dnorm(.x,mu,sigma)))
gaussian_df %>% head(3)
```

```{r}
colours <- c("Gaussian pdf"="red", "Binomial pmf"="blue")
fill <- c("Gaussian pdf"="white", "Binomial pmf"="white")
ggplot() + labs(x="x",y="Probability") + theme_bw() + geom_line(data=gaussian_df,aes(x=x,y=pdf,colour="Gaussian pdf"),size=2) +
geom_col(data=binom_df,aes(x=x,y=pmf,colour="Binomial pmf",fill="Binomial pmf")) + scale_color_manual(name = "", values=colours) +
scale_fill_manual(name = "", values=fill) + xlim(c(20,50))
```


## 5. Exponential distribution

```{r}
# Cumulative distribution function for an exponential distribution with parameter lambda
my_cdf_exp <- function(x,lambda){
  if (x<0) {return(0)}
  else {return(1-exp(-lambda*x))}
}
lambda <- 1/2
map_dbl(.x=seq(-1,4),.f=~my_cdf_exp(x=.x,lambda=lambda))
```

```{r}
# Confirming that my cdf exponential function works the same as the in-built function
test_inputs <- seq(-1,10,0.1)
my_cdf_output <- map_dbl(.x=test_inputs,.f=~my_cdf_exp(x=.x,lambda=lambda))
inbuilt_cdf_output <- map_dbl(.x=test_inputs,.f=~pexp(q=.x,rate=lambda))
all.equal(my_cdf_output,inbuilt_cdf_output)
```

```{r}
# Quantile function for an exponential distribution with parameter lambda
my_quantile_exp <- function(p,lambda){
  if (p<0 | p>1) {return("Invalid argument p, must have 0 <= p <= 1")}
  else {return(-ln(1-p)/lambda)}
}
map_dbl(.x=seq(0,1,0.2),.f=~my_quantile_exp(p=.x,lambda=lambda))
```

```{r}
# Confirming that my quantile exponential function works the same as the in-built function
test_inputs <- seq(0.01,0.99,0.01)
my_qf_output <- map_dbl(.x=test_inputs,.f=~my_quantile_exp(p=.x,lambda=lambda))
inbuilt_qf_output <- map_dbl(.x=test_inputs,.f=~qexp(p=.x,rate=lambda))
all.equal(my_qf_output,inbuilt_qf_output)
```


## 6. Poisson distribution

Answers written in notebook.