---
title: "SCEM Assignment 2"
author: "Max Kirwan"
date: "06/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Visualisation

```{r}
library(tidyverse)
library(Stat2Data)
data("Hawks")
hawksSmall<-drop_na(select(Hawks,Age,Day,Month,Year,CaptureTime,Species,Wing,Weight,Tail))
```

#### 1.1 Types of Variables

```{r}
dim(hawksSmall)
head(hawksSmall)
```

1. Month - Discrete
2. Species - Categorical
3. Age - Categorical
4. Wing - Continuous
5. Weight - Continuous

#### 1.2 What's wrong with this plot?

- No axis labels.
- No title.
- Too many visual cues... position, size and colour.

#### 1.3 Generate a histogram

```{r}
ggplot(hawksSmall, aes(x=Weight)) + geom_histogram(binwidth=100) + xlab("Weight (g)") + ylab("Count")
```

The weights have a bimodal distribution with 2 distinct means, probably due to the distinct species of Hawk.

#### 1.4 Generate a density plot

```{r}
ggplot(hawksSmall, aes(x=Tail)) + geom_density() + xlab("Tail (mm)") + ylab("Density")
```
```{r}
ggplot(hawksSmall, aes(x=Tail)) + geom_density(adjust=0.5) + xlab("Tail (mm)") + ylab("Density")
```
```{r}
ggplot(hawksSmall, aes(x=Tail, colour=Species)) + geom_density() + xlab("Tail (mm)") + ylab("Density")
```
```{r}
ggplot(hawksSmall, aes(x=Tail, y=Species, colour=Species, fill=Species)) + geom_violin() + xlab("Tail (mm)") + ylab("Species")
```

#### 1.5 Scatter plots

There are 4 aesthetics present:

- Position to Tail
- Position to Weight
- Colour to Species
- Shape to Species

There are 3 glyphs present:

- Red circles
- Green triangles
- Blue squares

```{r}
ggplot(hawksSmall, aes(x=Tail, y=Weight, colour=Species, shape=Species)) + geom_point() + xlab("Tail (mm)") + ylab("Weight (g)")
```

#### 1.6 Trend lines and facet wraps

```{r}
ggplot(hawksSmall, aes(x=Tail, y=Weight, colour=Species)) + geom_point() + geom_smooth(method="lm") + facet_wrap(vars(Species)) + xlab("Tail (mm)") + ylab("Weight (g)")
```

The visual cues used in this plot are position, colour, angle and direction.
In all 3 species of hawk there is a positive correlation between weight and tail length.

## 2. Data Wrangling

#### 2.1 Select and filter functions

```{r}
hSF <- Hawks %>%
  filter(Species=="RT", Weight>=1000) %>%
  select(Wing, Weight, Tail)
head(hSF)
```

The dataframe hSF has 3 variables; Wing, Weight and Tail and 398 examples/observations/cases.

#### 2.2 The arrange function

```{r}
head(arrange(hSF, Wing))
```

#### 2.2 Join and rename functions

```{r}
species_code <- c("CH","RT", "SS")
species_name_full <- c("Cooper's", "Red-tailed", "Sharp-shinned")
species_names <- data.frame(species_code, species_name_full)
species_names
```
```{r}
hawksFullName <- Hawks %>%
  left_join(species_names, by=c("Species"="species_code")) %>%
  select(-Species) %>%
  rename(Species = species_name_full)
head(select(hawksFullName, Species, Wing, Weight))
```

#### 2.4 The mutate function

```{r}
hawksWithBMI <- Hawks %>%
  mutate(bird_BMI = 1000*Weight/Wing^2) %>%
  filter(bird_BMI<=100) %>%
  select(Species, bird_BMI) %>%
  arrange(desc(bird_BMI))
head(hawksWithBMI)
```
```{r}
ggplot(hawksWithBMI, aes(x=bird_BMI, y=Species, colour=Species, fill=Species)) + geom_violin() + xlab("Bird BMI") + ylab("Species")
```

#### 2.5 Summarise and group-by functions

```{r}
hawksFullName %>%
  group_by(Species) %>%
  summarise(num_rows=n(), mean_wing=mean(Wing,na.rm=1), median_wing=median(Wing,na.rm=1), trimmed_mean_wing=mean(Wing,na.rm=1,trim=0.1), mean_wing_tail_ratio=mean(Wing/Tail,na.rm=1))
```

```{r}
hawksFullName %>%
  group_by(Species) %>%
  select(Wing, Weight, Culmen, Hallux, Tail, StandardTail, Tarsus, Crop) %>%
  summarise(across(everything(),~sum(is.na(.x))))
```

## 3. Exploratory Data Analysis

#### 3.1 Combining location estimators with the summarise function

```{r}
hawksFullName %>%
  group_by(Species) %>%
  summarise(Wing_mean=mean(Wing,na.rm=1), Wing_t_mean=mean(Wing,na.rm=1,trim=0.1), Wing_med=median(Wing,na.rm=1), Weight_mean=mean(Weight,na.rm=1), Weight_t_mean=mean(Weight,na.rm=1,trim=0.1), Weight_med=median(Weight,na.rm=1))
```

#### 3.2 Location and dispersion estimators under linear transformations

$X_1,...,X_n$ has sample mean $\mu$.

Let $\tilde{X_i} = a \cdot X_i + b$ for $a,b\in \mathbb{R}$.

Then the sample mean of $\tilde{X_i}$ is $$\frac{\tilde{X_1}+...+\tilde{X_n}}{n} = \frac{a \cdot (\tilde{X_1}+...+\tilde{X_n) + n \cdot b}}{n} = a \cdot \mu + b.$$


$X_1,...,X_n$ has sample variance $S^2_X = \frac{\Sigma(X_i - \mu)^2}{n-1}$.

Then $\tilde{X_i}$ has sample variance $$\frac{\Sigma(\tilde{X_i} - \tilde{\mu})^2}{n-1} = \frac{\Sigma(a \cdot X_i + b - a \cdot \mu - b)^2}{n-1} = \frac{a^2\Sigma(X_i - \mu)^2}{n-1} = a^2 \cdot S^2_X.$$

Also, $\tilde{X_i}$ has sample standard deviation $a \cdot S_X$ where $S_X$ is the sample standard deviation of $X_i$.


#### 3.3 Robustness of location estimators

```{r}
hal<-Hawks$Hallux # Extract the vector of hallux lengths
hal<-hal[!is.na(hal)] # Remove any nans
outlier_val<-100
num_outliers<-10
corrupted_hal<-c(hal,rep(outlier_val,times=num_outliers))
mean(hal)
mean(corrupted_hal)

num_outliers_vect<-seq(0,1000)
means_vect<-c()
for(num_outliers in num_outliers_vect){
  corrupted_hal<-c(hal,rep(outlier_val,times=num_outliers))
  means_vect<-c(means_vect,mean(corrupted_hal))
}

medians_vect<-c()
for(num_outliers in num_outliers_vect){
  corrupted_hal<-c(hal,rep(outlier_val,times=num_outliers))
  medians_vect<-c(medians_vect,median(corrupted_hal))
}

t_means_vect<-c()
for(num_outliers in num_outliers_vect){
  corrupted_hal<-c(hal,rep(outlier_val,times=num_outliers))
  t_means_vect<-c(t_means_vect,mean(corrupted_hal,trim=0.1))
}

df_means_medians<-data.frame(num_outliers=num_outliers_vect, mean=means_vect,t_mean=t_means_vect,
median=medians_vect)

df_means_medians%>%
  pivot_longer(!num_outliers, names_to = "Estimator", values_to = "Value")%>% 
  ggplot(aes(x=num_outliers,color=Estimator,linetype=Estimator,y=Value))+ geom_line()+xlab("Number of outliers")
```


#### 3.4 Box plots and outliers

```{r}
ggplot(Hawks, aes(x=Species,y=Weight)) + geom_boxplot()
```

```{r}
# Creating a function which computes the number of outliers within a sample
outliers <- function(x){
  q25 <- quantile(x,0.25,na.rm=1)
  q75 <- quantile(x,0.75,na.rm=1)
  IQR <- q75 - q25
  return(x[((x<q25-1.5*IQR)|(x>q75+1.5*IQR))&!is.na(x)])
}
num_outliers <- function(x){return(length(outliers(x)))}

Hawks %>%
  group_by(Species) %>%
  summarise(num_outliers_weight=num_outliers(Weight))
```

